Multi-reference pangenome graphs

### 2021-12-21 manuscript/main.tex
reorder

    A pangenome graph is a sequence model that encodes the mutual alignment of many genomes~\citep{Garrison_2019_thesis,Eizenga_2020}.
    In the variation graph, $V = (N, E, P)$, nodes $N = n_1\ldots n_{|N|}$ contain sequences of DNA.
    Each node $n_i$ has an identifier $i$ and an implicit reverse complement $\bar{n_i}$, and a node strand $s$ corresponds to one such orientations.
    Edges $E = e_1\ldots e_{|E|}$ represent ordered pairs of node strands: $e_i = ( s_a, s_b )$.
    Paths $P = p_1\ldots p_{|P|}$ describe walks over node strands: $p_i = s_1 \ldots s_{|p_i|}$.

    A set of pairwise alignments imply a variation graph, however, to our knowledge, there is no contained method that can
    generate the variation graph or lossless string graph from large input alignments.
    To explore this, we developed an algorithm to do so that operates in external memory by conversion of the alignment
    set into an alignment graph and the subsequent use of this graph in the elaboration of the variation graph including paths representing the input sequences.
    The resulting graph is a lossless representation of the input and the alignments between them.
    To distinguish the approach from string graphs, which imply error correction, we call this variation graph induction model the \emph{squish graph}.

    \FIXME{To integrate to the text} The alignment graph formalizes the collection of alignments between genomes.
    In this graph, each base pair in the input genomes is represented by a node.
    Two kinds of edges connect these nodes.
    The first `precedence' edge type indicates successive pairs of base pairs in the input,
    linking together the input genomes within the alignment graph.
    The second “match” type of edge links nodes that alignment has asserted to be matches.
    The alignment graph may be converted into a basic sequence graph by condensing all sets of nodes that are transitively linked by match edges.
    By recording the relationship between sequence graph and alignment graph nodes, we are able to trace the paths of pangenome sequences through the
    graph, producing our target: a pangenome variation graph that completely describes both the input assemblies and their alignments (Fig.~\ref{fig:from_alignment_to_variation_graph}).

    \texttt{seqwish} implements a lossless conversion from pairwise alignments between sequences to a variation graph encoding the sequences and their alignments.
    As input, we typically take all-versus-all alignments, but the exact structure of the alignment set may be defined in an application specific way.
    \texttt{seqwish} uses a series of disk-backed sorts and passes over the alignment and sequence inputs to allow the graph to be constructed in low memory relative to the size of the input sequence set.
    Memory usage during construction and traversal is limited by the use of sorted disk-backed arrays and succinct rank/select dictionaries to record a queryable version of the graph.

@@

    As input, we have $Q$, which is a concatenation of the sequences from which we will build the graph.
    We build a compressed suffix array (CSA) mapping sequence names to offsets in $Q$, and also the inverse using a rank/select dictionary on a bitvector marking the starts of sequences in $Q$.
    This allows us to map between positions in the sequences of $Q$, which is the format in which alignment algorithms typically express alignments, and positions in $Q$ itself, which is the coordinate space we will use as a basis for the generation of our graph.
    We encode the set of input pairwise alignments between sequences in $Q$ as object $A$.
    Although these alignments tend to be represented using oriented interval pairs in $Q$, for simplicity and robustness to graph complexity, we describe $A$ as a set of pairs of bidirectional positions (sequence offsets and strands) $[1 \ldots |Q_1 \ldots Q_{|Q|}|]$ , such that $A = \{ (b_{q}, b_{r}), \ldots \}$.
    We sort $A$ by the first member ($b_{q}$) of each pair, ensuring that the entries in $A$ are ordered according to their order in $Q$.

>>>>

A pangenome graph is a sequence model that encodes the mutual alignment of many genomes \citep{Garrison_2019_thesis,Eizenga_2020}.
In the variation graph, $V = (N, E, P)$, nodes $N = n_1\ldots n_{|N|}$ contain sequences of DNA.
Each node $n_i$ has an identifier $i$ and an implicit reverse complement $\bar{n_i}$, and a node strand $s$ corresponds to one such orientations.
Edges $E = e_1\ldots e_{|E|}$ represent ordered pairs of node strands: $e_i = ( s_a, s_b )$.
Paths $P = p_1\ldots p_{|P|}$ describe walks over node strands: $p_i = s_1 \ldots s_{|p_i|}$.

A variation graph represents all pairwise alignments between its paths.
It follows that we can construct a variation graph from a set of pairwise alignments.
We do so in a process based on three distinct graph models.
(\textit{1}) We transform the alignments and input sequences into an \textit{alignment graph} \cite{Kehr_2014}.
(\textit{2}) We then contract match components in this graph into single nodes of a \textit{sequence graph}.
(\textit{3}) By recording the translation from alignment graph nodes to sequence graph nodes, we are able to walk the input sequences through the graph.
In doing so, we are able to derive edges and label the paths of a \textit{variation graph} which captures the pairwise alignments and all implied transitive relationships between them.

%Finally, we derive edges and label the paths of the output variation graph by

%A set of pairwise alignments imply a variation graph.

%We can see that pairs of aligned sequences would be represented as single

%, however, to our knowledge, there is no existing method that can generate the variation graph or lossless string graph from large input alignments with arbitrary properties.

%We first formalize the graph induction algorithm that we will implement

%To explore this,  algorithm to do so that operates in external memory by conversion of the alignment set into an alignment graph and the subsequent use of this graph in the elaboration of the variation graph including paths representing the input sequences.
%The resulting graph is a lossless representation of the input and the alignments between them.
%To distinguish the approach from string graphs, which imply error correction, we call the result of this variation graph induction model the \emph{squish graph}.

\subsection{Sequence index}

As input, we have $Q$, which is a concatenation of the sequences from which we will build the graph.
We build a compressed suffix array (CSA) mapping sequence names to offsets in $Q$, and also the inverse using a rank/select dictionary on a bitvector marking the starts of sequences in $Q$.
This allows us to map between positions in the sequences of $Q$, which is the format in which alignment algorithms typically express alignments, and positions in $Q$ itself, which is the coordinate space we will use as a basis for the generation of our graph.

\subsection{Alignment graph}

\FIXME{To integrate to the text} The alignment graph formalizes the collection of alignments between genomes.
In this graph, each base pair in the input genomes is represented by a node.
Two kinds of edges connect these nodes.
The first `precedence' edge type indicates successive pairs of base pairs in the input,
linking together the input genomes within the alignment graph.
The second “match” type of edge links nodes that alignment has asserted to be matches.

We encode the set of input pairwise alignments between sequences in $Q$ as object $A$.
Although these alignments tend to be represented using oriented interval pairs in $Q$, for simplicity and robustness to graph complexity, we describe $A$ as a set of pairs of bidirectional positions (sequence offsets and strands) $[1 \ldots |Q_1 \ldots Q_{|Q|}|]$ , such that $A = \{ (b_{q}, b_{r}), \ldots \}$.
We sort $A$ by the first member ($b_{q}$) of each pair, ensuring that the entries in $A$ are ordered according to their order in $Q$.

\subsection{Sequence graph}

The alignment graph may be converted into a basic sequence graph by condensing all sets of nodes that are transitively linked by match edges.

\subsection{Variation graph}

By recording the relationship between sequence graph and alignment graph nodes, we are able to trace the paths of pangenome sequences through the graph, producing our target: a pangenome variation graph that completely describes both the input assemblies and their alignments (Fig.~\ref{fig:from_alignment_to_variation_graph}).

Seqwish implements a lossless conversion from pairwise alignments between sequences to a variation graph encoding the sequences and their alignments.
As input, we typically take all-versus-all alignments, but the exact structure of the alignment set may be defined in an application specific way.
Seqwish uses a series of disk-backed sorts and passes over the alignment and sequence inputs to allow the graph to be constructed in low memory relative to the size of the input sequence set.
Memory usage during construction and traversal is limited by the use of sorted disk-backed arrays and succinct rank/select dictionaries to record a queryable version of the graph.

@@
